A Process for the Emulation of Comparative Oncology Trials with Real-world Evidence (ENCORE)

### 2024-11-28 manuscript/manuscript.qmd
first draft of introduction

Randomized controlled trials (RCTs) have been the gold standard for establishing the efficacy and safety of medical products. With the advent of the the 21^st^ Century Cures Act directive, the Food and Drug Administration (FDA) established a framework to increasingly include and consider real-world evidence (RWE) generated from routine-care health data such as electronic health records (EHR) to evaluate and contextualize the comparative safety and effectiveness of novel cancer therapies.[@purpura2022role] With 21% of all drug approvals, oncology was the disease area with the most FDA drug approvals in 2023[@senior2024fresh], and especially in the field of precision oncology, RWE has a large potential to complement evidence coming from RCTs. Examples comprise the assessment of effectiveness in patient populations that are underrepresented in RCTs, the construction of external control arms in single-arm trials where active recruitment may not be feasible or the use of real-world data (RWD) for biomarker discovery and label extensions among pan-tumor populations that harbor specific genomic and immunological signatures.

However, the validity and transportability of results derived betwen RWE studies and RCTs can depend on many factors and limitations include missing data, small sample sizes, data discontinuity [@lin2018identifying; @joshua2022longitudinal], rapid changes in temporal prescribing patterns and the inability to measure and emulate common eligibility criteria and prognostic factors.[@rider2024emulations] While there are examples of emulations of oncology trials[@rider2024emulations; @merola2023aetion; @merola2024calibrating], a systematic and scaled approach to emulate a diverse set of different oncology trials is necessary to gain confidence the accuracy of RWE studies and to provide an answer as to which questions can be validly answered with which non-interventional study designs and analysis methods, given the data that is available.

@@

The *Emulation of Comparative Oncology Trials with Real-world Evidence* (**ENCORE**) project aims to extend this work to the field of oncology which comes with its own unique challenges that are not necessarily comparable with learnings from other disease areas and which must be systematically explored and understood.

@@

>>>>

Randomized controlled trials (RCTs) have been the gold standard for establishing the efficacy and safety of medical products. With the advent of the the 21^st^ Century Cures Act directive, the Food and Drug Administration (FDA) established a framework to increasingly include and additionally consider real-world evidence (RWE) generated from routine-care health data such as electronic health records (EHR) to evaluate and contextualize the comparative safety and effectiveness of novel cancer therapies.[@purpura2022role] With 21% of all approvals, oncology was the disease area with the most FDA drug approvals in 2023\[@senior2024fresh\], and especially in the field of precision oncology, RWE has a large potential to complement evidence coming from RCTs. Examples comprise the assessment of effectiveness in patient populations that are underrepresented in RCTs, the construction of external control arms in single-arm trials where active recruitment may not be feasible or the use of real-world data (RWD) for biomarker discovery and label extensions among pan-tumor populations that harbor specific genomic and immunological signatures.

However, the validity and transportability of results derived betwen RWE studies and RCTs can depend on many factors and frequently references limitations include missing data, small sample sizes, data discontinuity [@Merola2022;@joshua2022longitudinal], rapid changes in guideline treatment patterns and the inability to measure and emulate common eligibility criteria and prognostic factors in RWD.[@rider2024emulations] While there are publsihed examples of emulations of oncology trials[@rider2024emulations;@merola2023aetion; @merola2024calibrating], a systematic and scaled approach to emulate a diverse set of different oncology trials in various heterogeneous databases is necessary to gain confidence in the accuracy of RWE studies and to provide an answer as to which questions can be validly answered with which non-interventional study designs and analysis methods.

@@

The *Emulation of Comparative Oncology Trials with Real-world Evidence* (**ENCORE**) project [@encoreFDA] aims to extend this work to the field of oncology which comes with its own unique set of challenges that are not necessarily comparable with previous learnings from other disease areas and which must be systematically explored and understood. Building on a process co-developed with the FDA through RCT DUPLICATE[@wang2023emulation], this expansion to oncology is going to emulate 12 randomized clinical trials using multiple EHR data sources. The process includes an emphasis on transparency with documented assessment of data fitness of the RWD source for each trial [@rivera2024oncology;@gatto2022structured] and conducting extensive sensitivity analyses to assess robustness of findings and trial eligibility criteria.

The objectives of this project are to develop state-of-the-art methodological approaches and apply those to create insights that may provide guidance on the potential use of RWE for regulatory science in oncology. This includes the systematic evaluation of the suitability of data in relation to the study design and statistical analysis by emulating 12 oncology trials across four cancers and assessing the agreement of treatment effect estimates between trial emulation and RCT.

In this process paper, we describe the design and process for the selection of the 12 oncology RCTs, the assessment of the database quality and selection, protocol development, study design and statistical analysis and final agreement metrics to evaluate the concordance between RCT and its emulation.

@@

Trial selection

Data feasibility

Protocol development

Application of the target trial emulation framework to study design and statistical analysis

Agreement metrics
